### 2016-06-29 abstract.tex
edits

  As a result, database systems (in practice) support, and often
  encourage, developers to use weaker alternatives.  These
  alternatives break the strong isolation guarantees offered by
  serializable transactions to permit greater concurrency.
  Unfortunately, the semantics of weak isolation is often poorly
  understood, and usually explained only informally in terms of
  low-level implementation artifacts.  Consequently, verifying
  high-level correctness properties in such environments is
  problematic.

>>>>

  As a result, database systems (in practice) support, and often
  encourage, developers to use weaker alternatives.  These
  alternatives break the strong isolation guarantees offered by
  serializable transactions to permit greater concurrency.
  Unfortunately, the semantics of weak isolation is poorly understood,
  and usually explained only informally in terms of low-level
  implementation artifacts.  Consequently, verifying high-level
  correctness properties in such environments is problematic.

### 2016-06-29 introduction.tex
edits

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} have devised and experimented
with a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.

@@

Recent results on reasoning about application invariants in the
presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
  redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
At a high level, reasoning about weak consistency involves reasoning
about the concurrent execution of strongly atomic and fully isolated
operations against different (potentially replicated), but consistent
states of the database or data store, where each operation locally
preserves application invariants. In contrast, reasoning about weak
isolation involves reasoning about the concurrent execution of
transactions, whose operations witness different states which,
although consistent individually, may not be obviously reconciled into
a consistent global view.  Intuitively, the latter problem reduces to
the former if all transactions contain a single
operation. Furthermore, weak isolation can exist independent of weak
consistency, as evident from the presence of weakly isolated
transactions on sequentially consistent relational databases; such
systems existed long before the advent of weakly consistent NoSQL
stores.  Thus, frameworks for reasoning about weak isolation will
necessarily have to generalize the reasoning frameworks developed for
weak consistency in new and important ways.

@@

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly isolated transactions, demonstrating that a general
    isolation semantics can be expressed as well-formedness constraints
    on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are parameterized with specific weak isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

\noindent Our results thus provide the first (to the best of our
knowledge) mechanism that precisely relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly isolated
(serializable) counterparts.

The remainder of the paper is organized as follows.  The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:proofrules}
formalizes the proof systemt that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly consistent data stores.  We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:casestudies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

>>>>

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented
with a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.

@@

Recent results on reasoning about application invariants in the
presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
  redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
At a high-level, reasoning about weak consistency involves reasoning
about the concurrent execution of strongly atomic and fully isolated
operations against different (potentially replicated), but consistent
states of the database or data store, where each operation locally
preserves application invariants. In contrast, reasoning about weak
isolation involves reasoning about the concurrent execution of
transactions, whose operations witness different states which,
although consistent individually, may not be obviously reconciled into
a consistent global view.  Intuitively, the latter problem reduces to
the former if all transactions contain a single
operation. Furthermore, weak isolation can exist independent of weak
consistency, as evident from the presence of weakly isolated
transactions on sequentially consistent relational databases; such
systems existed long before the advent of weakly consistent NoSQL
stores.  Thus, frameworks for reasoning about weak isolation will
necessarily have to generalize the reasoning frameworks developed for
weak consistency in new and important ways.

@@

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parameteric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak-isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

\noindent Our results thus provide the first (to the best of our
knowledge) mechanism that precisely relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

The remainder of the paper is organized as follows.  The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly consistent data stores.  We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:casestudies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

### 2016-06-29 motivation.tex
edits

In this section, we motivate our proof system for weak isolation via
an example written in \txnimp - a C-like imperative language equipped
with a \C{txn} lexical block defining a transaction scope. Each
\C{txn} block is associated with a single-quoted string in angle
braces that uniquely identifies the transaction. We use Hoare triple
notation to annotate programs with pre- and post- conditions.

@@

Consider an implementation of a banking application that admits
concurrent withdraw transactions on a checking account (\C{C}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect our expectations. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficent
condition to preserve invaraints. But, is {\sc ser} necessary?

@@

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-executions}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it.\footnote{For clarity, the effects of different
  transactions are shown in different colored backgrounds.} In the
first execution, transaction {\bf Wd1} reads the current balance
(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
transaction {\bf Wd2} executes and commits, writing the new balance
(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
balance of \C{k-a1}. The second execution describes a similar scenario
with {\bf Wd1} and {\bf Wd2} exchanging their roles.  Clearly, {\sc
  rc} is an excessively weak isolation level for this program because
it loses the updates of one transaction, resulting in the violation of
the post-condition.  A stronger isolation level that prevents lost
updates is required.  \iso{Snapshot Isolation} ({\sc
  si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} is not functionally the same as
  {\sc ser}, however; the differences in the two manifest in the \emph{absence}
  of any write-write conflicts.}   Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based
concurrency control, it is also more efficient, making it appropriate
for both {\bf Wd} transactions.

@@

Thinking in terms of anomalies, as described above, is how database
programmers are often encouraged to reason about weak isolation.
Unfortunately, such reasoning does not rest on any sound foundation,
and is thus highly error-prone. Reasoning in terms of how weak
isolation variants are implementated is no better since it requires
programmers to understand low-level implementation details
of the database that are naturally far removed from the application
semantics. An attractive (and principled) alternative in this context
is a formal proof system that combines declarative reasoning about
isolation guarantees with operational reasoning about programs. We
demonstrate how our proof system makes this possible in the context of
the current example.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behaviour of individual threads by
abstracting away interference induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under weak isolation, determining what
constitutes an interference is a non-trivial problem. For example,
inside a serializable transaction there is no interference, whereas
within a transaction executing under a \iso{Read Uncommitted}
isolation level, the weakest isolation level, all interferences are
valid. Between these two extremes are various levels of isolation that
admit some interferences while prohibiting others.  For example,
\iso{Read Committed} isolation admits interference of a committed
transaction, but not that of an uncommitted transaction.
\iso{Snapshot Isolation} admits interference from committed
non-conflicting transactions, and so on.

@@

The specification of an isolation level encodes its interference
characteristic as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

@@

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc sc} to totally
order all operations w.r.t the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different store
consistency semantics, supporting stores that are weaker than {\sc sc}
(e.g., eventually consistent), and should be able to reconcile
conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

>>>>

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope.  For exposition convenience, each
\C{txn} block is associated with a single-quoted string in angle
braces that uniquely identifies the transaction. We use Hoare triple
notation to annotate programs with pre- and post- conditions.

@@

Consider an implementation of a banking application that admits
concurrent withdraw transactions on a checking account (\C{C}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficent
condition to preserve invaraints. But, is {\sc ser} necessary?

@@

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-executions}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it.\footnote{For clarity, the effects of different
  transactions are shown in different colored backgrounds.} In the
first execution, transaction {\bf Wd1} reads the current balance
(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
transaction {\bf Wd2} executes and commits, writing the new balance
(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
balance of \C{k-a1}. The second execution describes a similar scenario
with {\bf Wd1} and {\bf Wd2} exchanging their roles.  Clearly, {\sc
  rc} is an excessively weak isolation level for this program because
it loses the updates of one transaction, resulting in the violation of
the post-condition.  A stronger isolation level that prevents lost
updates is required.  \iso{Snapshot Isolation} ({\sc
  si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} exhibits different behaviour from
  {\sc ser} in the absence of write-write conflicts.}   Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based
concurrency control, it is also more efficient, making it appropriate
for both {\bf Wd} transactions.

@@

Thinking in terms of anomalies, as described above, is how database
programmers are often encouraged to reason about weak isolation.
Unfortunately, such reasoning does not rest on any sound foundation,
and is thus highly error-prone. Reasoning in terms of how weak
isolation variants are implementated is no better since it requires
programmers to understand low-level implementation details of the
database that are naturally far removed from application semantics. An
attractive (and principled) alternative in this context would be a
formal proof system that combines declarative reasoning about
isolation guarantees with operational reasoning about programs. We
demonstrate how our proof system makes this possible in the context of
the current example.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behaviour of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under weak isolation, determining what
constitutes an interference is a non-trivial problem. For example,
inside a serializable transaction there is no interference, whereas
within a transaction executing under a \iso{Read Uncommitted}
isolation level, the weakest isolation level, all interferences are
valid. Between these two extremes are various levels of isolation that
admit some interferences while prohibiting others.  For example,
\iso{Read Committed} isolation admits interference of a committed
transaction, but not that of an uncommitted transaction.
\iso{Snapshot Isolation} admits interference from committed
non-conflicting transactions, and so on.

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

@@

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc sc} to totally
order all operations w.r.t the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different store
consistency semantics, supporting stores that are weaker than {\sc sc}
(e.g., causally or eventually consistent), and should be able to
reconcile conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

### 2016-06-29 opsem.tex
edits

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

@@

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visiblity
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operation semantics for isolation levels, solely by relying on their
declarative characterization as trace well-formedness constraints.
\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
isolation levels stated as trace well-formedness constraints.

>>>>

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp\
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

@@

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visiblity
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operationalsemantics for isolation levels, instead solely relying on
their declarative characterization as trace well-formedness
constraints to enforce isolation guarantees.
\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
isolation levels stated as trace well-formedness constraints.
